Using raw accelerometer data to predict high-impact mechanical loading

### 2022-02-27 manuscript/manuscript.Rmd
Add objective
Write the introduction

---
# title:
output:
  bookdown::pdf_document2:
    toc: false
    includes:
      in_header: "preamble.tex"
    latex_engine: lualatex
  word_document:
    reference_docx: style_reference.docx
always_allow_html: true
bibliography: bibliography.bib
csl: nlm.csl
---

@@

@@

After familiarization, participants performed a series of trials envolving jumps of different types and heights. First, drop jumps were executed from steps varying from 5cm to 40cm high, with 5cm increments. Then, box jumps were performed from the floor to 5cm, 15cm and 30cm high boxes. Finally, continuous jumps were carried out in two different heights: 5cm and 15cm. Drop and box jumps were executed in two sets of four jumps at each height, with 30s interval, and the continuous jumps in one set of 20s at each height. The landing phase of the drop jumps, the take-off phase of the box jumps and both phases of the continuous jumps were recorded in force plates (1000Hz; AMTI Corporation, Watertown, USA). Participants were instructed to jump with no restrictions to arms movement, land two-footed and then remain stationary on the force plate for five seconds (except in the continuous jumps).

@@

The MATLAB software (Version 2019a, Mathworks, Natick, USA) was used to process data from both FP and accelerometers. Data were processed according to the following procedures: first, the GRF signal was resampled to match the activity monitors sampling frequency (100Hz). Then, both GRF and ACC signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, to attenuate the noise magnitude. This cut-off frequency was selected based on visual inspection of the signals frequency spectrum after a Fast Fourier Transform (FFT). Afterwards, the GRF and ACC resultant vector was calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, both signals were adjusted using the time set by the systems clock and then synchronized by the maximum cross-correlation coefficient, with posterior manual adjustments if necessary. Next, peak ACC (pACC) were determined by the following criteria: i) a minimum height of three standard deviations above the mean ACC value recorded during a trial; and ii) a separation of at least 0.2 seconds for continuous jumps and 4 seconds for the other jumps. Peak GRF (pGRF) were then defined as the highest value between 0.2 or 4 seconds before and after each pACC, according to the jump type. The rates of change were computed through a centered derivative from the beginning of the foot contact to the curve peak (Equation 1). The peak loading rate (pLR) and peak acceleration rate (pAR) were defined as the maximum value in the array (Equation 2).

>>>>

---
title: "Using raw accelerometer data to predict high-impact mechanical loading"
output:
  bookdown::pdf_document2:
    toc: false
    includes:
      in_header: "preamble.tex"
    latex_engine: lualatex
  word_document:
    reference_docx: style_reference.docx
always_allow_html: true
bibliography: bibliography.bib
csl: nlm.csl
---

@@

# Introduction {.unnumbered}

Bone reponds to mechanical loading by altering its properties, such as density, geometry, and strength [@Turner2003; @Fonseca2014; @Santos2017]. Physical activity is a way of eliciting mechanical stimuli, and it is well established to improve bone health. However, there is no consensus in which characteristics of the physical activity-related mechanical stimulus are better associated with bone health benefits. Nevertheless, there is evidence that that high-impact exercise, although performed scarcely throughout the day, may play an important role in inducing positive bone adaptations [@Vainionpaa2006; @Stiles2017].

But, even when knowing precisely which is the best loading regimen to improve bone health, the question of how to measure and monitor this load remains. Bone mechanical loading is usually measured in laboratory settings, using instruments such as force plates (FP) [@Medved2000]. Measurements in this instruments, that are confined to a laboratory space, are not applicable to controling the loads a person is routinely subjected in free-living contexts. To surpass this issue, in the last few years, methods using wearable devices such as accelerometers have been proposed to measure mechanical loading, more precisely,  ground reaction force (GRF) [@Neugebauer2012; @Neugebauer2014; @Fortune2014; @Neugebauer2018; @Veras2020] and loading rate (LR) [@Fortune2014; @Veras2020]. These devices are light, small, can be used throughout days, and have a relatively affordable price. The aforementioned characteristics make the accelerometers well suited instruments to measure mechanical loading on a daily basis.

However, the validity of the accelerometer-predicted mechanical loading values depends on some factors, as the sample characteristics [@Welk2005]. In fact, due to this, several calibration studies have been done in several populations, as children [@Neugebauer2012], adults [@Neugebauer2014; @Fortune2014], military personnel [@Neugebauer2018], and subjects with obesity [@Veras2020]. Another important factor that influence the predictions validity is the type of activities performed in the calibration study [@Bassett2012]. Until now, studies that calibrate accelerometers to predict mechanical loading variables have employed usually activities as walking [@Fortune2014; @Neugebauer2018; @Veras2020], or walking and running [@Neugebauer2012; @Neugebauer2014]. Nevertheless, activities eliciting higher impact loads have not yet been tested.

Therefore, the objective of this study was to develop accelerometry-based mechanical loading prediction equations for high-impact activites in subjects from normal weight to severe obesity.

@@

In this study, jumps were chosen as the high-impact activites to be performed. After familiarization, participants completed a series of trials envolving jumps of different types and heights. First, drop jumps were executed from steps varying from 5cm to 40cm high, with 5cm increments. Then, box jumps were performed from the floor to 5cm, 15cm and 30cm high boxes. Finally, continuous jumps were carried out in two different heights: 5cm and 15cm. Drop and box jumps were executed in two sets of four jumps at each height, with 30s interval, and the continuous jumps in one set of 20s at each height. The landing phase of the drop jumps, the take-off phase of the box jumps and both phases of the continuous jumps were recorded in FC (1000Hz; AMTI Corporation, Watertown, USA). Participants were instructed to jump with no restrictions to arms movement, land two-footed and then remain stationary on the force plate for five seconds (except in the continuous jumps).

@@

The MATLAB software (Version 2019a, Mathworks, Natick, USA) was used to process data from both FP and accelerometers. Data were processed according to the following procedures: first, the GRF signal was resampled to match the activity monitors sampling frequency (100Hz). Then, both GRF and ACC signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, to attenuate the noise magnitude. This cut-off frequency was selected based on visual inspection of the signals frequency spectrum after a Fast Fourier Transform (FFT). Afterwards, the GRF and ACC resultant vector was calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, both signals were adjusted using the time set by the systems clock and then synchronized by the maximum cross-correlation coefficient, with posterior manual adjustments if necessary. Next, peak ACC (pACC) were determined by the following criteria: i) a minimum height of three standard deviations above the mean ACC value recorded during a trial; and ii) a separation of at least 0.2 seconds for continuous jumps and 4 seconds for the other jumps. Peak GRF (pGRF) were then defined as the highest value between 0.2 or 4 seconds before and after each pACC, according to the jump type. The rates of change were computed through a centered derivative from the beginning of the foot contact to the curve peak (Equation 1). The peak LR (pLR) and peak acceleration rate (pAR) were defined as the maximum value in the array (Equation 2).
